---
title: "p8105_hw3_ys3924"
author: "YISU"
date: "2024-10-12"
output: github_document
---

```{r setup, include=FALSE}
library(p8105.datasets)
library(tidyverse)
library(ggplot2)
library(patchwork)
```

# Problem 1

### Read in and clean NY NOAA data
```{r}
data("ny_noaa") 
ny_noaa = ny_noaa |>
  janitor::clean_names() |> 
  separate(date, c('year', 'month', 'day'), sep ='-') |>
  mutate(prcp = as.numeric(prcp),
         tmax = as.numeric(tmax),
         tmin = as.numeric(tmin),
         year = as.numeric(year),
         prcp = prcp / 10,
         tmax = tmax / 10,
         tmin = tmin / 10)
ny_noaa_row = nrow(ny_noaa)
```


The dataset contains key variables id, date, prcp, snow, tmin, tmax. The dataset has `r ny_noaa_row` observations. There exists missing values for temperature, precipitation, or snowfall, which can affect analysis if not handled properly.


Count the snow fall
```{r}
snowfall_counts = ny_noaa |>
  count(snow) |>
  arrange(desc(n)) |>
  slice(1)
snowfall_counts
```

For snowfall, the most commonly observed values is 0, which contains 2008508 observations.



### Create plot showing the average max temperature in January and in July in each station across years

```{r, message=FALSE}
# Filter the data
jan_jul_temp = ny_noaa |>
  filter(month %in% c("01", "07")) |>
  group_by(id, year, month) |>
  drop_na(tmax) |>
  summarise(avg_max_temp = mean(tmax, na.rm = TRUE)) |>
  ungroup()

# Create the plot using the filtered data
  ggplot(jan_jul_temp, aes(x = year, y = avg_max_temp, color = id)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~month, labeller = labeller(month = c("01" = "January", "07" = "July"))) +
  labs(title = "Average Max Temperature of Stations in January and July across Years and Stations",
       x = "Year",  
       y = "Average Max Temperature") + 
  theme_minimal() +
  theme(legend.position = "none")
```

According to the above graph, average max temperature in July is higher than in January, and temperatures appear to be more consistent during July. The outliers exist for both January and July, where temperatures drop significantly below the typical range for most stations.


### Create plot showing tmax vs tmin and distribution of snowfall values 
```{r, warning=FALSE}
# Plot for tmax vs tmin  for the full dataset
plot_i = ggplot(ny_noaa, aes(x = tmin, y = tmax)) +
  geom_hex() +
  labs(title = "tmax vs tmin for the full Dataset", x = "Min Temperature", y = "Max Temperature") +
  theme_minimal()
# Plot for distribution of snowfall values
plot_ii = ny_noaa |>
  filter(snow > 0 & snow < 100) |>
  group_by(year) |>
  summarise(avg_snow = mean(snow, na.rm = TRUE)) |>
  ggplot(aes(x = year, y = avg_snow)) +
  geom_col(width = 0.7) +
  labs(title = "Snowfall Distribution by Year", x = "Year", y = "Average Snowfall") +
  theme_minimal()
plot_i + plot_ii
```


# Problem 2

### Read in and clean accelerometer and demographic data
```{r, message=FALSE}
accel_data = read_csv("./data/nhanes_accel.csv") |>
  janitor::clean_names() |>
  mutate(seqn = as.character(seqn))
covar_data = read_csv("./data/nhanes_covar.csv", skip = 3) |>
  janitor::row_to_names(row_number = 1) |>
  janitor::clean_names() |>
  mutate(
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
    education = factor(education, levels = c(1, 2, 3), 
                       labels = c("Less than high school",
                                  "High school equivalent",
                                  "More than high school")),
    age = as.numeric(age)
  ) |>
  drop_na() |>
  filter(age >= 21)
full_data = left_join(covar_data, accel_data, by = "seqn")
```

### Create table for the number of men and women in each education category
```{r, message=FALSE}
full_data |>
  group_by(sex, education) |>
  summarise(count = n()) |>
  pivot_wider(names_from = education, values_from = count) |>
  knitr::kable()
```

By looking at above table, there are slightly more women have achieved higher levels of education (more than high school) compared to men, while more men tend to fall in the "High school equivalent" category. For education level less than high school, the balance between men and women is fairly even.

### Create a visualization of the age distributions for men and women in each education category

```{r}
ggplot(full_data, aes(x = education, y = age, fill = sex)) +
  geom_boxplot() +
  labs(title = "Age Distribution by Gender and Education Category", 
       x = "Education Category", 
       y = "Age") +
  theme_minimal()
```

According to above boxplot, the age distribution across education levels shows that individuals with less education, particularly "Less than high school", tend to  be slightly older. The "High school equivalent" category shows a younger age range for men than women. For the "More than high school" category, the distribution are similar, except that the median age for men is sligtly higher than women.


### Plot total activities against age

```{r, message=FALSE}
full_data = full_data |>
  mutate(total_activity = rowSums(across(starts_with("min")),na.rm = TRUE))
  
ggplot(full_data, aes(x = age, y = total_activity, colour = sex)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~education) +
  labs(title = "Total Activity vs. Age by Gender and Education Level",
       x = "Age",
       y = "Total Activity") +
  theme_minimal()
```

From above plots, total activity of males and females decrease as age increase across all educational level. For "less than high school", the total activity for women is higher before age of 40, after which men's total activity starts to be larger. For "high school equivalent", the total activity for both men and women increases, but after age of 40 it decreases sharply. For "More than high school", the women's total activity is higher than men's for all age.


### Create plot for 24-hour activity time courses for each education level 

```{r, warning=FALSE}
full_data |>
  pivot_longer(cols = starts_with("min"), names_to = "minute", values_to = "activity") |>
  mutate(minute = as.numeric(str_remove(minute, "min"))) |>
  group_by(minute, sex, education) |>
  summarize(avg_activity = mean(activity)) |>
  ggplot(aes(x = minute, y = avg_activity, color = sex)) +
  geom_line(alpha = 0.2) +
  geom_smooth() +
  facet_wrap(~ education) +
  labs(title = "24-Hour Activity Time Courses by Gender for Education Level", 
       x = "Minute",
       y = "Average Activity") +
  theme_minimal()
```

By looking at above plot, activity peaks occur in the middle of the day and gradually decline afterward across all education levels, which follows the human daily routine. Women tend to have higher activity levels than men across most of the day in all education levels, particularly in "More than high school" group. For "Less than high school" group, activity peaks around midday and gradually decreases. For "High school equivalent" group, the activity pattern is more stable. For "More than high school" group, it shows a slightly higher plateau compared to the other groups.


# Problem 3

### Read in and clean Citi Bike data

```{r, message=FALSE}
jan_2020 = read_csv("./data/citibike/Jan 2020 Citi.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  mutate(year = 2020,
         month = "January")
jan_2024 = read_csv("./data/citibike/Jan 2024 Citi.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  mutate(year = 2024,
         month = "January")
july_2020 = read_csv("./data/citibike/July 2020 Citi.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  mutate(year = 2020,
         month = "July")
july_2024 = read_csv("./data/citibike/July 2024 Citi.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  mutate(year = 2024,
         month = "July")
citi_data = bind_rows(jan_2020, jan_2024, july_2020, july_2024)
citi_rows = nrow(citi_data)
citi_cols = ncol(citi_data)
```

The cleaned and merged dataset contains `r citi_cols` variables and `r citi_rows` observations. The key variables include rideable_type, ride_id, weekdays, duration, start_station_name, end_station_name, member_casual, year, and month.

### Produce table showing the total number of rides

```{r}
citi_data |>
  group_by(year, month, member_casual) |>
  summarise(total_rides = n()) |>
  pivot_wider(
    names_from = "member_casual",
    values_from = "total_rides") |>
  knitr::kable()
```
Across all time periods, Citi Bike members consistently outnumber casual riders in terms of total rides. Also, both rider types show an increase in rides from January to July in each year. Over the span of four years, the number of both casual and member riders grew from 2020 to 2024.




